The very dependent recursive structure of iterated parametricity in indexed form

### 2025-12-29 paper/paper.tex
paper: minor corrections

For $\nu$ an arity, that is $1$ in the case of predicates, families and augmented semi-simplicial sets or $2$ in the case of relations, correspondences and semi-cubical sets, such sequence of was called a $\nu$-set. Equations specifying $\nu$-sets were given in \cite{hr25}. Because of the dependency of $E_{n+1}$ into all $E_p$ for $p \leq n$, the family could not be expressed as a function. Instead, it was expressed as a stream whose $(n+1)$-th element $E_{n+1}$ was dependent on the bundle of all previous elements of the stream. Such a bundle was called $\Xto{m}{n}$ while the residual stream of all $E_{n}$, $E_{n+1}$, ..., all of them precisely dependent on the bundle $E:\Xto{m}{n}$ of previous $E_0$, ..., $E_{n-1}$, was called $\Xfrom{m}{n}{D}$. The type of $E_n$, also dependent on a bundle $D:\Xto{m}{n}$, was denoted $\Xcomp{m}{n}[D=\D]$, as shown on tables~\ref{tab:coind} and~\ref{tab:core}, using an type theory as language of discourse. When $n$ is $0$, $\Xto{m}{0}$ is degenerated, that is reduced to the canonical inhabitant $\kstar$ of $\unittype$. Then, $\Xfrom{m}{0}{\kstar}$ denotes the full stream $E_0, E_1, \ldots$.

@@

To build $\framep[n+1]$ from $\frame[n]$, it happens to be enough to assume the following built at level $n$:

\begin{equation*}
  \begin{array}{lclr}
    \framep[n][p]   & : & \U                                                                                                  & \mbox{for all $p\leq n$} \\
    \painting[n][p] & : & \framep[n][p] \rightarrow \U                                                                        & \mbox{for all $p\leq n$} \\
    \restrf[n][p]   & : & \Pi q \leq n-p.\, \Pi \epsilon.\, \framep[n+1][p] \rightarrow \framep[n][p]                         & \mbox{for all $p\leq n$} \\
    \restrp[n][p]   & : & \Pi q \leq n-p.\, \Pi \epsilon.\, \Pi d:\framep[n+1][p].\,                                                                     \\
                    &   & \painting[n+1][p](d) \rightarrow \painting[n][p](\restrf[n][p](d))                                  & \mbox{for all $p\leq n$} \\
    \cohf[n][p]     & : & \Pi qr\epsilon\omega.\, \restrf[n][p][q][\epsilon] \circ \restrf[n+1][p][\omega][r] =                                          \\
                    &   & \restrf[n][p][\omega][r] \circ \restrf[n+1][p][q+1][\epsilon]                                       & \mbox{for all $p\leq n$} \\
    \cohp[n][p]     & : & \Pi qr\epsilon\omega.\,\Pi d:\framep[n+2][p].\,                                                                                \\
                    &   & \restrp[n][p][q][\epsilon](\restrf[n+1][p][r][\omega](d)) \circ \restrp[n+1][p][r][\omega](d) =                                \\
                    &   & \restrp[n][p][r][\omega](\restrf[n+1][p][q+1][\epsilon](d)) \circ \restrp[n+1][p][q+1][\epsilon](d) & \mbox{for all $p\leq n$} \\
  \end{array}
\end{equation*}
and to build it at level $n+1$. This is however not so easy because some components of this structure, e.g. $\restrf$ and further fields, depend on $\framep[n+1]$, that is already depend on the ability of building the $\framep$ component of the structure at level $n+1$. We are thus facing an interdependency between specifying the structure assumed at level $n$ and defining from it the structure at level $n+1$. So, we have to proceed in order: specifying $\framep[n]$, from which $\painting[n]$ can be specified, from which $\framep[n+1]$ can be built and thus $\restrf[n+1]$ be specified. Now, the definition of $\framep[n+1][p]$, which is by induction on $p$, itself depends on $\restrf[n][p-1]$ whose type depends in turn on $\framep[n+1][p-1]$. This means that $\framep[n+1][p]$ and the type of $\restrf[n][p]$ have to be mutually defined.

The definition of $\framep[n+1][p]$ also depends, via $\layer[n][p-1]$, on $\painting[n][p-1]$. Both via $\painting[n][p-1]$ and $\restrf[n][p-1]$, it depends also on $\framep[n][p-1]$ by typing dependency. This requires to assume $\framep[n][p-1]$ and $\painting[n][p-1]$ known in advance. Moreover, the definition of $\framep[n+1][p]$ depends on $\framep[n+1][p-1]$, so all of $\framep[n][k]$, $\painting[n][k]$ and $\restrf[n][k]$ for $k< p$ need to be known, which we respectively abbreviate as $\framep[n][\lbrack 0,p-1 \rbrack]$, $\painting[n][\lbrack 0,p-1 \rbrack]$ and $\restrf[n][\lbrack 0,p-1 \rbrack]$. Schematically, using dotted arrows for dependencies in the type and plain arrows for dependencies in the definition, this gives the following dependencies for $p\leq n+1$ fixed:

@@

So far, using this recursive definition of  $\framep[n][\lbrack 0,p-1 \rbrack]$ and $\restrf[n][\lbrack 0,p-1 \rbrack]$, we are able to specify the three first component of the exected dependent structure:
\begin{equation*}
  \begin{array}{lclr}
    \framep[n][p]             & :                        & \U                                                & \mbox{for all $p\leq n$} \\
    \painting[n][p]           & :                        & \framep[n][p] \rightarrow \U                      & \mbox{for all $p\leq n$} \\
    \restrf[n][p]             & :                        & \Pi q \leq n-p.\, \Pi \epsilon.\, \framep[n+1][p]
    \left(\begin{array}{l}
              \framep[n][\lbrack 0,p-1 \rbrack]   \\
              \painting[n][\lbrack 0,p-1 \rbrack] \\
              \restrf[n][\lbrack 0,p-1 \rbrack]   \\
            \end{array}\right)
    \rightarrow \framep[n][p] & \mbox{for all $p\leq n$}                                                                                \\
  \end{array}
\end{equation*}
which can appropriately be reformulated using sequences for appropriate definitions of $\frametype[n][\lbrack 0,p \rbrack]$, $\paintingtype[n][\lbrack 0,p \rbrack]$ and $\restrftype[n][\lbrack 0,p \rbrack]$, since the dependencies of $\restrf[n][p]$ into $\restrf[n][\lbrack 0,p-1 \rbrack]$ makes it a very dependent function~\citep{Hickey96}:
\begin{equation*}
  \begin{array}{lcl}
    \framep[n][\lbrack 0,p \rbrack]   & : & \frametype[n][\lbrack 0,p \rbrack]                                     \\
    \painting[n][\lbrack 0,p \rbrack] & : & \paintingtype[n][\lbrack 0,p \rbrack](\framep[n][\lbrack 0,p \rbrack]) \\
    \restrf[n][\lbrack 0,p \rbrack]   & : & \restrftype[n][\lbrack 0,p \rbrack]
    \left(\begin{array}{l}
              \framep[n][\lbrack 0,p-1 \rbrack]   \\
              \painting[n][\lbrack 0,p-1 \rbrack] \\
              \restrf[n][\lbrack 0,p-1 \rbrack]   \\
            \end{array}\right)                                                                      \\
  \end{array}
\end{equation*}

>>>>

For $\nu$ an arity, that is $1$ in the case of predicates, families and augmented semi-simplicial sets, or $2$ in the case of relations, correspondences and semi-cubical sets, such sequence of was called a $\nu$-set. Equations specifying $\nu$-sets were given in \cite{hr25}. Because of the dependency of $E_{n+1}$ into all $E_p$ for $p \leq n$, the family could not be expressed as a function. Instead, it was expressed as a stream whose $(n+1)$-th element $E_{n+1}$ was dependent on the bundle of all previous elements of the stream. Such a bundle was called $\Xto{m}{n}$ while the residual stream of all $E_{n}$, $E_{n+1}$, ..., all of them precisely dependent on the bundle $E:\Xto{m}{n}$ of previous $E_0$, ..., $E_{n-1}$, was called $\Xfrom{m}{n}{D}$. The type of $E_n$, also dependent on a bundle $D:\Xto{m}{n}$, was denoted $\Xcomp{m}{n}[D=\D]$, as shown on tables~\ref{tab:coind} and~\ref{tab:core}, using an type theory as language of discourse. When $n$ is $0$, $\Xto{m}{0}$ is degenerated, that is reduced to the canonical inhabitant $\kstar$ of $\unittype$. Then, $\Xfrom{m}{0}{\kstar}$ denotes the full stream $E_0, E_1, \ldots$.

@@

To build $\framep[n+1]$ from $\framep[n]$, it happens to be enough to assume the following built at level $n$, for all $p \leq n$:

\begin{equation*}
  \begin{array}{lcl}
    \framep[n][p]   & : & \U                                                                                                  \\
    \painting[n][p] & : & \framep[n][p] \rightarrow \U                                                                        \\
    \restrf[n][p]   & : & \Pi q \leq n-p.\, \Pi \epsilon.\, \framep[n+1][p] \rightarrow \framep[n][p]                         \\
    \restrp[n][p]   & : & \Pi q \leq n-p.\, \Pi \epsilon.\, \Pi d:\framep[n+1][p].\,                                          \\
                    &   & \painting[n+1][p](d) \rightarrow \painting[n][p](\restrf[n][p](d))                                  \\
    \cohf[n][p]     & : & \Pi qr\epsilon\omega.\, \restrf[n][p][q][\epsilon] \circ \restrf[n+1][p][\omega][r] =               \\
                    &   & \restrf[n][p][\omega][r] \circ \restrf[n+1][p][q+1][\epsilon]                                       \\
    \cohp[n][p]     & : & \Pi qr\epsilon\omega.\,\Pi d:\framep[n+2][p].\,                                                     \\
                    &   & \restrp[n][p][q][\epsilon](\restrf[n+1][p][r][\omega](d)) \circ \restrp[n+1][p][r][\omega](d) =     \\
                    &   & \restrp[n][p][r][\omega](\restrf[n+1][p][q+1][\epsilon](d)) \circ \restrp[n+1][p][q+1][\epsilon](d) \\
  \end{array}
\end{equation*}
and to build it at level $n+1$. This is however not so easy because some components of this structure, e.g. $\restrf$ and further fields, depend on $\framep[n+1]$, that is already depend on the ability of building the $\framep$ component of the structure at level $n+1$. We are thus facing an interdependency between specifying the structure assumed at level $n$ and defining from it the structure at level $n+1$. So, we have to proceed in order: specifying $\framep[n]$, from which $\painting[n]$ can be specified, from which $\framep[n+1]$ can be built and thus $\restrf[n+1]$ be specified. Now, the definition of $\framep[n+1][p]$, which is by induction on $p$, itself depends on $\restrf[n][p-1]$ whose type depends in turn on $\framep[n+1][p-1]$. This means that $\framep[n+1][p]$ and the type of $\restrf[n][p]$ actually have to be mutually defined.

The definition of $\framep[n+1][p]$ also depends, via $\layer[n][p-1]$, on $\painting[n][p-1]$. Both via $\painting[n][p-1]$ and $\restrf[n][p-1]$, it depends also on $\framep[n][p-1]$ by typing dependency. This requires to assume $\framep[n][p-1]$ and $\painting[n][p-1]$ known in advance. Moreover, the definition of $\framep[n+1][p]$ depends on $\framep[n+1][p-1]$, so all of $\framep[n][k]$, $\painting[n][k]$ and $\restrf[n][k]$ for $k< p$ need to be known, which we respectively abbreviate as $\framep[n][\lbrack 0,p-1 \rbrack]$, $\painting[n][\lbrack 0,p-1 \rbrack]$ and $\restrf[n][\lbrack 0,p-1 \rbrack]$. Schematically, using dotted arrows for dependencies in the type and plain arrows for dependencies in the definition, this gives the following graph of dependencies for $p\leq n+1$ fixed:

@@

So far, using this recursive definition of  $\framep[n][\lbrack 0,p-1 \rbrack]$ and $\restrf[n][\lbrack 0,p-1 \rbrack]$, we are able to specify the three first component of the expected dependent structure, for $p \leq n$:
\begin{equation*}
  \begin{array}{lcl}
    \framep[n][p]   & : & \U                                                \\
    \painting[n][p] & : & \framep[n][p] \rightarrow \U                      \\
    \restrf[n][p]   & : & \Pi q \leq n-p.\, \Pi \epsilon.\, \framep[n+1][p]
    \left(\begin{array}{l}
              \framep[n][\lbrack 0,p-1 \rbrack]   \\
              \painting[n][\lbrack 0,p-1 \rbrack] \\
              \restrf[n][\lbrack 0,p-1 \rbrack]   \\
            \end{array}\right)
    \rightarrow \framep[n][p]                                               \\
  \end{array}
\end{equation*}
which can appropriately be reformulated using sequences for appropriate definitions of $\frametype[n][\lbrack 0,p \rbrack]$, $\paintingtype[n][\lbrack 0,p \rbrack]$ and $\restrftype[n][\lbrack 0,p \rbrack]$, since the dependencies of $\restrf[n][p]$ into $\restrf[n][\lbrack 0,p-1 \rbrack]$ makes it a very dependent function~\citep{Hickey96}:
\begin{equation*}
  \begin{array}{lcl}
    \framep[n][\lbrack 0,p \rbrack]   & : & \frametype[n][\lbrack 0,p \rbrack]                                     \\
    \painting[n][\lbrack 0,p \rbrack] & : & \paintingtype[n][\lbrack 0,p \rbrack](\framep[n][\lbrack 0,p \rbrack]) \\
    \restrf[n][\lbrack 0,p \rbrack]   & : & \restrftype[n][\lbrack 0,p \rbrack]
    \left(\begin{array}{l}
              \framep[n][\lbrack 0,p-1 \rbrack]   \\
              \painting[n][\lbrack 0,p-1 \rbrack] \\
              \restrf[n][\lbrack 0,p-1 \rbrack]   \\
            \end{array}\right)                                                                      \\
  \end{array}
\end{equation*}
